---
title: "Data Analysis Using Base R"
author: "Ilyas Ustun"
output: 
    html_document:
        toc: true
        toc_float: 
            collapsed: false
            smooth_scroll: true
---

twitter: @_ilyas_ustun_


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this lesson we will learn the basics of data analysis in base R.

R has built-in datasets which can be checked with `data()`.

```{r}
data()

mtcars
```

Check the structure of the data
```{r}
str(mtcars)
```

Create a summary of each variable (column) using `summary()`
```{r}
summary(mtcars)
```

The outcome of summary is a `table`.
```{r}
sm = summary(mtcars)
class(sm)
sm[1, ]
```

It is a good idea to check the first few and last few rows.   
With `str()`, `summary()`, `tail()`, and `head()` you can get an overall understanding of the data.  
This will help you spot problems in the data.  

```{r}
head(mtcars)
```

```{r}
tail(mtcars)
```

To select one of the variables we can use:  
- `$` symbol after the dataframe name  
- Selecting by index integer  
- Selecting by boolean vector  
- Selecting by column names  

But, be careful! The returned variables can be of different types (vector, dataframe)  

Using the `$` symbol:
```{r}
mtcars$cyl
```

Selecting by index integer:
```{r}
mtcars[2]
```

```{r}
mtcars[,2]
```


Selecting by boolean vector:
```{r}
colnames(mtcars)
mask = "cyl" == colnames(mtcars)
mask
```

```{r}
mtcars[mask]
```

```{r}
mtcars[,mask]
```


Selecting by column names:
```{r}
mtcars['cyl']
```

```{r}
mtcars[, 'cyl']
```


Selecting by column names:
`[[]]` returns a vector
```{r}
mtcars[['cyl']]
```


For categorical variables, or numeric variables with a few unique values we can get the counts table using `table()`.

Let's first check how many unique values the cylinder variable has:
```{r}
unique(mtcars$cyl)
```


Let's see how many vehicles there are in each of the cylinder type: 
```{r}
table(mtcars$cyl)
```


Creating a crosstab:
```{r}
tab = table(mtcars$cyl, mtcars$gear)
tab
rownames(tab) = paste(rownames(tab), 'cyl') 
tab
```



## Filtering rows

```{r}
head(mtcars)
```

Only retain those rows where the number of cylinders is equal to 6:
```{r}
mtcars$cyl == 6
```


```{r}
mtcars[mtcars$cyl == 6, ]
```
Only retain those rows where the number of cylinders is **NOT** equal to 6:
```{r}
mtcars[mtcars$cyl != 6, ]
```

## Filtering on several conditions

Only retain those rows where the number of cylinders (`cyl`) is **NOT** equal to 6, **AND** the `hp` is greater than 150:
```{r}
mtcars[(mtcars$cyl != 6) & (mtcars$hp > 150), ]
```


## Filtering (Rows) and Selection (Columns)

Retain the rows where `cyl` is equal to 6, and only keep the columns `cyl`, `wt`, `mpg`. 
```{r}
mtcars[mtcars$cyl == 6, c('cyl', 'wt', 'mpg')]
```

Retain the rows where `cyl` is equal to 6 OR `wt` is smaller than 3, and only keep the columns `cyl`, `wt`, `mpg`. 
```{r}
mtcars[(mtcars$cyl == 6) | (mtcars$wt < 3), c('cyl', 'wt', 'mpg')]
```

```{r}
mtcars[(mtcars$cyl == 6) | (mtcars$wt < 3), colnames(mtcars) %in% c('cyl', 'wt', 'mpg')]
```

## Creating summary statistics

```{r}
summary(mtcars)
```

What is the average mpg for each cylinder type?
```{r}
mtcars$mpg[mtcars$cyl == 4]
mean(mtcars$mpg[mtcars$cyl == 4], na.rm=TRUE)
```

```{r}
mean(mtcars$mpg[mtcars$cyl == 6], na.rm=TRUE)
```

```{r}
mean(mtcars$mpg[mtcars$cyl == 8], na.rm=TRUE)
```

```{r}
tapply(mtcars$mpg, mtcars$cyl, mean)
```



Thank you!  
Workshop organized at DSI 2022  
November 19, 2022  
Houston, TX    

By: Ilyas Ustun  
Jarvis College of Computing  
DePaul University  
Chicago, IL  
iustun@depaul.edu    

 
Twitter:  
@_ilyas_ustun_   


