---
title: "Data Wrangling and Analysis Using dplyr in R"
author: "Ilyas Ustun"
output: 
    html_document:
        toc: true
        toc_float: 
            collapsed: true
            smooth_scroll: true
---


twitter: @_ilyas_ustun_


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this lesson, we will enhance our data analysis and wrangling capabilities by using a popular package called `dplyr`. This package that contains many data wrangling and data manipulation functions (also called verbs). One of the benefits of dplyr is its consistency in its outputs, compare that to base R in which you can get a vector, dataframe or even a list as output for the same activities.

First we will install `dplyr` library. 
```{r}
# install the dplyr library
# install.packages("dplyr")

# Make the library available in the current session
library(dplyr)
```

Install the `nycflights13` library which contains the flights data originating from New York airports in 2013.
```{r}
# install.packages("nycflights13")

# Make the library available in the current session
library(nycflights13)
```

```{r}
head(flights)
class(flights)
```



## Selecting columns (including)

Select the `origin` column.
```{r}
select(flights, origin) %>% head(10)
```

Select the `origin` and the `dest` columns.
```{r}
select(flights, origin, dest) %>% head(10)
```


## Exclude columns
We can exclude columns by putting a `-` (minus) to the left of them.
```{r}
select(flights, -flight, -tailnum, -carrier) %>% head()
```

```{r}
select(flights, -c(flight, tailnum, carrier)) %>% head()
```

## Filtering rows
Rows can be filtered based on conditions.  

Retain the rows where carrier is equal to "DL".
```{r}
filter(flights, carrier == 'DL') %>% head()
```

### "OR" operator '|'

Retain the rows where carrier is equal to "DL" **OR** the destination is equal to "ORD".

```{r}
df = filter(flights, carrier == 'DL' | dest == "ORD")

# verify that this is true
table(df$dest, df$carrier)
```

When combining AND and OR statements, we need to be careful about ordering and use parenthesis to state order of computation.

Carrier is DL or destination is ORD. Then we want to filter those rows where month is equal to 5 (May).
```{r}
filter(flights, ((carrier == 'DL') | (dest == "ORD")) & (month == 5))
```

To make this condition easier to understand we can divide it into two, and call two filter statements one after another. 
```{r}
flights %>%
  filter(carrier == 'DL' | dest == "ORD") %>%
  filter(month == 5)
```

Filter the flights data where departure delay and arrival delay are both greater than 500 minutes.
```{r}
filter(flights, dep_delay > 500, arr_delay > 500)
```


Ex: Create a subset of data from flights where:  
- the features `carrier`, `dest` (destination airport), and `dep_delay` (departure delay) are selected.  
- departure delay is more than 100 minutes  
- destination airport is ORD  

How many rows and columns are there?

```{r}
flights %>%
    select(carrier, dest, dep_delay) %>%
    filter(dep_delay > 100, dest == 'ORD') %>%
    glimpse()
```



Ex: Create a subset of data from flights where:  
- the features `carrier`, `dest` (destination airport), and `dep_delay` (departure delay) are selected.  
- departure delay is more than 100 minutes  
- destination airport is ORD  
- arrival delay is more than 50 minutes

How many rows and columns are there?

If you try to do something with a column that wasn't selected you will get an error

```{r}
# flights %>%
#   select(carrier, dest, dep_delay) %>%
#   filter(dep_delay > 100, dest == 'ORD', arr_delay > 50)
#   
```


```{r}
flights %>%
    filter(dep_delay > 100, dest == 'ORD', arr_delay > 50) %>%
    select(carrier, dest, dep_delay) %>%
    str()
```


Ex: Create a subset of data from flights dataset where:  
The features selected are:   
    - carrier
    - destination airport
    - departure delay
Conditions:  
    - departure delay more than 50 minutes
    - destination ORD or LAX

How many rows are there in this dataset?

```{r}
flights %>%
  select(carrier, dest, dep_delay) %>%
  filter(dep_delay > 50) %>%
  filter(dest == 'ORD' | dest == 'LAX') %>% str()
```


Ex. Create a subset of data from flights dataset where
The features selected are:
    - carrier
    - destination airport
    - departure delay
Conditions:
    - departure delay more than 50 minutes
    - destination ORD, LAX, SFO, BOS, MIA, ATL

How many rows are there in this dataset?


```{r}
flights %>%
    select(carrier, dest, dep_delay) %>%
    filter(dep_delay > 50) %>%
    filter(dest %in% c("ORD", "LAX", "SFO", "BOS", "MIA", "ATL")) %>%
    glimpse()
```

## Slicing

Get rows with certain index value

We want to filter rows with index: 1,3,4 (1st, 3rd, and 4th rows)

```{r}
flights %>%
  slice(c(1,3,4))
```

```{r}
flights %>% head()
```


## Sorting

Sorts the dataframe based on values of specified variables

Let's sort by the values of departure delay
```{r}
flights %>%
    arrange(dep_delay) %>%
    head(10)
```

If there are repeating values in the first feature selected for ordering, we can select a second feature to order those rows.
```{r}
flights %>%
    arrange(dep_delay, arr_delay) %>% 
    head(10)
```


For character variables the ordering is based on alphabetical order.
```{r}
flights %>%
    select(origin, dest, flight) %>% 
    head(20) %>%
    arrange(dest)
```


Descending order:
```{r}
set.seed(123) # Set the seed to make the process repeatable
flights %>%
    select(origin, dest, flight) %>% 
    sample_n(20) %>% # Select 20 random points before ordering to make it faster
    arrange(desc(dest))
```


## Creating a new variable

### Create a new variable using base R:

Create total delay by adding departure delay and arrival delay.
```{r}
flights$total_delay = flights$dep_delay + flights$arr_delay
flights %>% str()
```



### Create a new variable using `mutate` in dplyr:

```{r}
flights = select(flights, -total_delay)

flights = flights %>%
    mutate(total_delay = arr_delay + dep_delay)

str(flights)
```

Create speed by making use of distance and `air_time`.
```{r}
flights = flights %>%
    mutate(speed = distance/ (air_time/60))

str(flights)
```

Ex. What is the maximum speed in all of the flights?
Order in descending speed.
```{r}
flights %>%
    arrange(desc(speed)) %>%
    head(10)
```


```{r}
flights %>%
  arrange(desc(total_delay),desc(speed))
```


Ex. What is the maximum speed among the airplanes that experienced a total delay of more than 100 minutes but less than 600 minutes?

```{r}
flights %>% 
    select(total_delay, speed) %>%
    filter(total_delay > 100, total_delay < 600) %>%
    arrange(desc(speed)) %>%
    # mutate(speed = round(speed)) %>%
    slice(1)
```

Ex. How many flights had a speed greater than 500 mph and experienced a total delay of more than 200 minutes but less than 400 minutes?

```{r}
flights %>%
  filter(speed > 500, total_delay > 200, total_delay < 400) %>%
  count()
```



## Summarizing Data

Get the mean value of total delay variable.
```{r}

mean(flights$total_delay, na.rm=TRUE)

flights %>% summarize(avg_tot_delay = mean(total_delay, na.rm=TRUE),
                      median_tot_delay = median(total_delay, na.rm=TRUE),
                      max_speed = max(speed, na.rm=TRUE),
                      n_rows = n())
```
## Grouping

Slice - calculate - put together
Group by month, and then summarize for each month value
```{r}
flights %>%
    group_by(month) %>%
    summarize(N = n(), 
            avg_tot_delay = mean(total_delay, na.rm=TRUE),
            avg_arr_delay = mean(arr_delay, na.rm=TRUE),
            avg_dep_delay = mean(dep_delay, na.rm=TRUE))
```

## Quiz:
### Q1:
1. Group by carrier
2. How much of average total delay is experienced by each carrier?
3. What is the standard deviation of the total delay by each carrier?
4. How many flights were operated by each carrier?

```{r}
flights %>%
    group_by(carrier) %>% #P1
    summarize( 
            avg_tot_delay = mean(total_delay, na.rm=TRUE), # P2
            sd_tot_delay = sd(total_delay, na.rm=TRUE), #P3
            N = n()) %>% # P4
    arrange(-avg_tot_delay)
```


### Q2:
1. Group by carrier and month
2. How much of average total delay is experienced by each carrier in the month of July?
3. What is the standard deviation of the total delay by each carrier in the month of May?
4. How many flights were operated by each carrier in December?
5. Which airline had the largest average speed in January?


#### Q2. General solution
```{r}
q2 = flights %>%
    select(carrier, month, total_delay, speed) %>% 
    group_by(carrier, month) %>%
    summarize(avg_tot_delay = mean(total_delay, na.rm=TRUE),
              sd_total_delay = sd(total_delay, na.rm = TRUE),
              N = n(),
              avg_speed = mean(speed, na.rm=TRUE)) %>%
    filter(month %in% c(1, 5, 7, 12)) %>%
    ungroup()

str(q2)
```

#### Solution Q2-P2)
```{r}
q2 %>% filter(month == 7) %>% arrange(-avg_tot_delay)
```

#### Solution Q2-P3)
```{r}
q2 %>% filter(month == 7) %>% arrange(-sd_total_delay)
```



### Q3:
Find the worst destination airport in terms of average arrival delay in the month of December?
(worst means here largest average arrival delay)


```{r}
flights %>%
  filter(month==12) %>%
  group_by(dest) %>%
  summarize(avg_arr_delay = mean(arr_delay, na.rm=TRUE)) %>%
  arrange(desc(avg_arr_delay))
```

### Q4:
Find the worst destination airport for the flights originated from LGA airport in terms of average arrival delay in the month of December?
(worst means here largest average arrival delay)



```{r}
flights %>%
  filter(month==12, origin=='LGA') %>%
  group_by(dest) %>%
  summarize(avg_arr_delay = mean(arr_delay, na.rm=TRUE)) %>%
  arrange(desc(avg_arr_delay))
```


Thank you!  
Workshop organized at DSI 2022  
November 19, 2022  
Houston, TX    

By: Ilyas Ustun  
Jarvis College of Computing  
DePaul University  
Chicago, IL  
iustun@depaul.edu    

 
Twitter:  
@_ilyas_ustun_   











